Statistical analysis of proteomics experiments with R and MSstats (v3.0.9)

### 2015-09-16 MSstats_manual_RMarkdown.Rmd
Correct few typos and make changes in some sentences
More to be made as soon as I go through the source code once again.

Setting the working directory to where you saved files.

```{r, eval=T}
setwd('/Users/Meena/Dropbox/MSstats_GitHub_document')
getwd()
```

If you didn't have MSstats installed so far. Please install it now.

@@

Load MSstats and verify that you have the correct version (3.0.9) loaded.

@@

## 1. preparing the data for MSstats input

@@

```{r, eval=T}
DDA2009.superhirn<-read.csv("RawData.DDA.csv")

@@

The default option for normalization is `equalizeMedians`, constant normalization shifts all the intensities in a run by a constant, to equalize the median of intensities across runs for label free experiment. This normalization works when we can assume the majority of proteins do not change across proteins. Therefore if your label free DDA dataset has small number of proteins, please be careful to use `equalizeMedians` Instead, if you have a spiked in standard then you want to set this to `globalStandards` and define the standard with `nameStandards`
For label based experiment, `equalizeMedians` equalizes the median of reference intensities across runs for label based experiment. therefore, it works for dataset with small number of proteins.

Usually label-free DDA datasets have many missing values and noisy features with outliers. MSstats supports several ways to deal with this.  The default option for summarization is `TMP` (Tukey's median polish, robust parameter estimation method with median across rows and columns) after imputation by AFT (accelerated failure time model, `MBimpute=TRUE`) based on censored intensity for NA (`censoredInt="NA"`, which assumes that NA intensities is censored, not detected due to below the cutoff) with cutoff value for censored = minimum of minimum value for a feature and minimum value for a run (`cutoffCensored="minFeatureNRun"`). It can handle missing values by imputing and outliers from TMP estimation. However, the runs, which have no measurements at all, will be removed and not be used for any calculation.

```{r, eval=T}
# default option
DDA2009.TMP <- dataProcess(raw = DDA2009.superhirn,  fillIncompleteRows = TRUE,
                           normalization = 'equalizeMedians',
                           summaryMethod = 'TMP',
                           censoredInt="NA", cutoffCensored="minFeatureNRun",
                           MBimpute=TRUE)
```

If 'MBimpute=FALSE', imputation based on AFT model will be not performed and censored intensities (here NAs) will be just replace with cutoff value assigned in `cutoffCensored="minFeatureNRun"`. Then the runs with more than 50% measurement missing will use cutoff value, which is biased by cutoff value. In this case, you can remove the runs ths have more than 50 % missing values in analysis with `remove50missing=TRUE`.

@@

 If 'censoredInt=NULL', we assume that all intensities are randomly missing and there is no action for missing values.

@@

Besides summarizing observations with linear models MSstats also offers a summarization option using linear model, and as sum of log-intensities, which is the default Skyline behaviour. `summaryMethod="linear"` with `censoredInt=NULL` assumes that all NAs are randomly missing and use `lm` or `lmer` for parameter estimation. `summaryMethod="linear"` with `censoredInt="NA"` assumes that NA intensities are censored, not detected because intensities are below the cutoff. Then use AFT model with left-censored for parameter estimation. 'cutoffCensored' is the same as `summaryMethod="TMP"`. For `summaryMethod="linear"`, AFT model is only used for parameter estimation, not for imputation. Therefore `MBimpute=TRUE` with `summaryMethod="linear"` doesn't work (no imputation before parameter estimation, the same with `MBimpute=FALSE`).

```{r, eval=F, message=F, warning=F}
# linear mixed model (lm or lmer) with run and feature
DDA2009.linear <- dataProcess(raw = DDA2009.superhirn, 
                              summaryMethod="linear", censoredInt=NULL)

# accerated failure model with left-censored. NA intensities are assumed as censored
DDA2009.linear.censored <- dataProcess(raw = DDA2009.superhirn,
                                       summaryMethod="linear", censoredInt="NA")
```

@@

Now let's look at what the equalize medians procedure did to our data. We can generate these for all proteins but also for single Proteins at a time if we have a big dataset. Let's look at the 'Kininogen-1' (Kng1/NP_036828) protein in these example data.

@@

Let's look at how the original data look like. Profile plot show all individual measurements per protein. With type="profileplot", two pdfs will be generated. The first pdf includes plots (per protein) to show individual measurement for each peptide (peptide for DDA, transition for SRM or DIA) across runs, grouped per condition. Ech peptide has a different colour/type layout. If you dont want to generate these plots, please use 'originalPlot=FALSE'. The second pdf, which named with 'wSummarization' suffix, show run-level summarized data per protein. The same peptides (or transition) in the first plot are presented in grey, with the summarized values by TMP overlayed in red.

@@

Have a look at the profile plots to compare the different options (TMP with considering missing values vs. TMP without considering missing values for summarizing in dataProcess steps.) Original profile plots are the same. But, summarization plots are different, especially for condition 'C1' and 'C2' in 'yeast' protein, which have many missing values.

@@

After we normalized the data and summarized each protein's behaviour across conditions with one of the dataProcess summarization methods, we are all set to compare protein changes between groups of conditions. Within MSstats we can do this with the `groupComparison` function, which takes as input the output of the `dataProcess` function.

@@

Of course we have to tell `groupComparison` which are the conditions we would like to compare...  
We define our contrast matrix by adding a column for every condition, **in alphabetical order**.

@@

It shows the order of condition, which MSstats can recognize. Then, We add a row for every comparison we would like to make between groups of conditions.

@@

For example, if you want to compare C2-C1, which means log(C2)-log(C1) and the same as log(C2/C1), set '1' for C2 and '-1' for C1 in the row. Also combination of multiple groups are possible. For example, if you want to compare between average of C2 and C3 and average of C1, (C3+C2)/2-C1 as formula, set '-1' for C1, '0.5' for C2 and '0.5' for C3, and '0' for rest of groups.

@@

Let's inspect the results to see what proteins are changing significantly between groups.

```{r, eval=TRUE}
# output from groupComparison function has three data.frame
names(DDA2009.comparisons)

# name of columns in result data.frame
names(DDA2009.comparisons$ComparisonResult)

# get only siginificant proteins and comparions among all comparisons
SignificantProteins = 
  DDA2009.comparisons$ComparisonResult[DDA2009.comparisons$ComparisonResult$adj.pvalue < 0.05 ,]
nrow(SignificantProteins)
```

@@

Volcano plots visualize the outcome of one comparison between conditions for all the pro-
teins, and combines the information on statistical and practical significance. The y-axis displays
the FDR-adjusted p-values on the negative log2 scale, and represents statistical significance.
The horizontal dashed line represents the FDR cutoff. The points above the FDR cutoff line
are statistically significant differentially abundant proteins. These points are colored in red for
upregulated proteins, and in blue for downregulated proteins. The x-axis is the model-based
estimate of log-fold change (the base of logarithm transform is the same as specified in the
logTrans option of the dataProcess step), and represents practical significance. It is possible
to specify a practical significance cutoff based on the estimate of fold change in addition to
the statistical significance cutoff. If the fold change cutoff is specified, the points above the
horizontal cutoff line but within the vertical cutoff line will be judged as not differentially
abundant (and will be colored in black). The practical significance cutoff can only be applied
in addition to the statistical significance cutoff (i.e. the fold change alone does not present
enough evidence for differential abundance).

@@

Then, 'VolcanoPlot.pdf' will be saved under the folder you assigned. It has the plots per comparison you set in contrast.matrix. Below is one of volcano plot, for comparison 'C6-C5'. Please check `?groupComparisonPlots' for detail, such as labelling protein names, size of dots, font sizes, etc.

@@

Heatmaps illustrate the patterns of up- and down-regulation of proteins in several comparisons. Columns in the heatmaps are comparison of conditions you assigned in contrast.matrix, and rows are proteins. The heatmaps display signed FDR-adjusted p-values of the tests, colored in red/blue for significantly up-/down-regulated proteins, while taking into account the specified FDR cutoff and the additional optional fold change cutoff. Brighter colors
indicate stronger evidence in favor of differential abundance. Black color represents proteins
are not significantly differentially abundant.

@@

## 1. preparing the data for MSstats input
Let's read the data.

```{r, eval=T}
raw<-read.csv(file="ControlMixerMSstatsInputfromskyline.csv")
head(raw)
```

The raw data (input data for MSstats) is required to contain variable of ProteinName, PeptideSequence, PrecursorCharge, FragmentIon, ProductCharge, IsotopeLabelType, Condition, BioReplicate, Run, Intensity. The variable names should be fixed. MSstats input from Skyline adapts the column scheme of the dataset so that it fits MSstats input format. However you should change `FileName` to `Run` and `Area` to `Intensity`.

@@

The difference between output from Skyline and other spectral processing tool is that Skyline can distinguish random missing (NA) and censored missing (zero). The output from Skyline can have NA (expect small number of NAs or none of them) and 0(zero). Then, we can use zero values in intensity are censored.

@@

### 1.1 preprocessing with DDA experiment from Skyline output

@@

### 2. different option for dataProcess
Only different options in `dataProcess` you should use for Skyline output are `skylineReport=TRUE` (remove `Truncated=TRUE` rows and handle intensity=0) and `censoredInt="0"`(use intensity=0 as censored to handle missing values).

@@

If you have MaxQuant output, please follow this section. Here controlled mixture data with dynamic range bench mark (Cox, 2014 in MCP) is used to demonstrate.

### 1. First, get protein ID information
```{r, eval=T}
proteinGroups<-read.table("DDA2014_proteinGroups.txt", sep="\t", header=TRUE)
```

### 2. Read in annotation including condition and biological replicates: annotation.csv
```{r, eval=T}
annot <- read.csv("DDA2014_annotation.csv", header=TRUE)
```

### 3. Read in MaxQuant file: evidence.txt
```{r, eval=F}
infile <- read.table("evidence.txt", sep="\t", header=TRUE)
```

```{r, eval=T, echo=FALSE}
load("DDA2014.evidence.RData")
infile<-DDA2014.evidence
```

@@

```{r, eval=T}
msstats.raw<-MaxQtoMSstatsFormat(evidence=infile, annotation=annot, proteinGroups=proteinGroups)

>>>>

Set the working directory to where you saved files. Note that you should a different path on your computer from the example.

```{r, eval=T}
setwd('/Users/Meena/Dropbox/MSstats_GitHub_document')
```

You can check your working directory by:

```{r, eval=T}
getwd()
```

If you didn't have MSstats installed so far, you can downlaod the package from the [MSstats installation page](http://msstats.org/installation/) to your working directory and install it as follows:

@@

Load MSstats into the R session and verify that you have the correct version (3.0.9) loaded.

@@

## 1. Preparing the data for MSstats input

@@

```{r, eval=T}
DDA2009.superhirn <- read.csv("RawData.DDA.csv")

@@

The default option for normalization is `equalizeMedians`, where all the intensities in a run are shifted by a constant, to equalize the median of intensities across runs for label-free experiment. This normalization works when the assumption that the majority of proteins do not change across runs is held. Therefore if your label-free DDA dataset has a small number of proteins, please be careful to use the `equalizeMedians` option. Instead, if you have a spiked in standard then you want to set this to `globalStandards` and define the standard with `nameStandards`. For label based experiment, `equalizeMedians` equalizes the median of reference intensities across runs for label based experiment. Therefore, it works for dataset with a small number of proteins.

Usually label-free DDA datasets have many missing values and noisy features with outliers. MSstats supports several ways to deal with this. The default option for summarization is `TMP` (Tukey's median polish, robust parameter estimation method with median across rows and columns) after imputation by AFT (accelerated failure time model, `MBimpute=TRUE`) based on censored intensity for NA (`censoredInt="NA"`) with a cutoff as the minimum value for a feature and a run (`cutoffCensored="minFeatureNRun"`). Here the NA values represent censored measurements, which are below the cutoff and not detected. This process handles missing values through imputation and reduces the influence of the outliers using the TMP estimation. However, those runs with no measurements at all will be removed and not be used for any calculation.

```{r, eval=T}
# default option
DDA2009.TMP <- dataProcess(raw = DDA2009.superhirn,  fillIncompleteRows = TRUE,
                           normalization = 'equalizeMedians',
                           summaryMethod = 'TMP',
                           censoredInt = "NA", cutoffCensored="minFeatureNRun",
                           MBimpute = TRUE)
```

If `MBimpute=FALSE`, imputation will not be performed and censored intensities (here NAs) will be just replaced with cutoff value assigned in `cutoffCensored="minFeatureNRun"`. With this option, those runs with substantial missing measurements will be biased by the cutoff value. In such case, you can remove the runs that have more than 50% missing values from your analysis with `remove50missing=TRUE`.

@@

If `censoredInt=NULL`, we assume that all intensities are randomly missing and there is no action for missing values.

@@

Besides summarizing observations with the median polish method, MSstats also offers a summarization option using linear model, and one based on the sum of log-intensities, which is the default Skyline behaviour. Summarization using the option `summaryMethod="linear"` with `censoredInt=NULL` assumes that all NAs are randomly missing and uses `lm` or `lmer` for parameter estimation. On the other hand, `summaryMethod="linear"` with `censoredInt="NA"` assumes that NA intensities are censored, not detected because intensities are below the cutoff. AFT model is then applied with left-censored for parameter estimation. 'cutoffCensored' is the same as `summaryMethod="TMP"`. For `summaryMethod="linear"`, AFT model is only used for parameter estimation, not for imputation. Therefore `MBimpute=TRUE` with `summaryMethod="linear"` doesn't work (no imputation before parameter estimation, the same with `MBimpute=FALSE`).

```{r, eval=F, message=F, warning=F}
# linear mixed model (lm or lmer) with run and feature
DDA2009.linear <- dataProcess(raw = DDA2009.superhirn, 
                              summaryMethod = "linear", censoredInt = NULL)

# accerated failure model with left-censored. NA intensities are assumed as censored
DDA2009.linear.censored <- dataProcess(raw = DDA2009.superhirn,
                                       summaryMethod = "linear", censoredInt = "NA")
```

@@

Now let's look at what the equalize medians procedure did to our data. We can generate QC plots for all proteins or for single proteins at a time if we have a big dataset. Let's look at the 'Kininogen-1' (Kng1/NP_036828) protein in these example data.

@@

Let's look at how the original data look like. Profile plot shows all individual measurements per protein. With `type="profileplot"`, two pdfs will be generated. The first pdf includes plots (per protein) to show individual measurement for each peptide (peptide for DDA, transition for SRM or DIA) across runs, grouped per condition. Each peptide has a different color/type layout. If you don't want to generate these plots, please use the option `originalPlot=FALSE`. The second pdf, which is named with 'wSummarization' suffix, shows run-level summarized data per protein. The same peptides (or transition) in the first plot are presented in grey, with the summarized values by TMP overlaid in red.

@@

Have a look at the profile plots to compare the different options (e.g., TMP with or without considering missing values for summarization in `dataProcess`.) Original profile plots are the same. But, summarization plots are different, especially for conditions 'C1' and 'C2' in 'yeast' protein, which have many missing values.

@@

After we normalized the data and summarized each protein's behaviour across conditions with one of the `dataProcess` summarization methods, we are all set to compare protein changes between groups of conditions. Within MSstats we can do this with the `groupComparison` function, which takes as input the output of the `dataProcess` function.

@@

We need to provide the `groupComparison` function with a contrast matrix to define the comparison to be made. The contrast matrix is created with each condition in column and each comparison in row. Note that the conditions are arranged **in alphabetical order**:

@@

It shows the order of condition, which MSstats can recognize. Then, we add a row with entries of 0, 1, or -1 for every comparison we would like to make between groups of conditions.

@@

For example, if you want to compare C2-C1, which means log(C2)-log(C1) and the same as log(C2/C1), set '1' for C2 and '-1' for C1 in the row. Also, combining multiple groups for comparison is possible. For example, if you want to compare between average of C2 and C3 and average of C1, (C3+C2)/2-C1 as formula, set '-1' for C1, '0.5' for C2 and '0.5' for C3, and '0' for rest of groups.

@@

Let's inspect the results to see what proteins change significantly between groups.

```{r, eval=TRUE}
# output from groupComparison function has three data frames
names(DDA2009.comparisons)
```

Results of the statistical comparison are stored in the data frame named `ComparisonResult`:

```{r, eval=TRUE}
# name of columns in result data.frame
names(DDA2009.comparisons$ComparisonResult) 
```

You can get a shorter list of the differentially abundant proteins based on the comparison results:

```{r, eval=TRUE}
# get only significant proteins and comparisons among all comparisons
SignificantProteins = 
  DDA2009.comparisons$ComparisonResult[DDA2009.comparisons$ComparisonResult$adj.pvalue < 0.05 ,]
nrow(SignificantProteins)
```

@@

Volcano plots visualize the outcome of one comparison between conditions for all the proteins, and combine the information on statistical and practical significance. The y-axis displays
the FDR-adjusted p-values on the negative log10 scale, and represents statistical significance.
The horizontal dashed line represents the FDR cutoff. The points above the FDR cutoff line
are statistically significant differentially abundant proteins. These points are colored in red for
upregulated proteins, and in blue for downregulated proteins. The x-axis is the model-based
estimate of log-fold change (the base of logarithm transform is the same as specified in the
logTrans option of the dataProcess step), and represents practical significance. It is possible
to specify a practical significance cutoff based on the estimate of fold change in addition to
the statistical significance cutoff. If the fold change cutoff is specified, the points above the
horizontal cutoff line but within the vertical cutoff line will be judged as not differentially
abundant (and will be colored in black). The practical significance cutoff can only be applied
in addition to the statistical significance cutoff (i.e. the fold change alone does not present
enough evidence for differential abundance).

@@

Then, 'VolcanoPlot.pdf' will be saved under the folder you assigned. It has the plots per comparison you set in contrast.matrix. Below is one of volcano plot, for comparison 'C6-C5'. Please check `?groupComparisonPlots` for detail, such as labelling protein names, size of dots, font sizes, etc.

@@

Heatmaps illustrate the patterns of up- and down-regulation of proteins in several comparisons. Columns in the heatmaps are comparison of conditions you assigned in `contrast.matrix`, and rows are proteins. The heatmaps display signed FDR-adjusted p-values of the tests, colored in red/blue for significantly up-/down-regulated proteins, while taking into account the specified FDR cutoff and the additional optional fold change cutoff. Brighter colors
indicate stronger evidence in favor of differential abundance. Black color represents proteins
are not significantly differentially abundant.

@@

## 1. Preparing the data for MSstats input
Let's read the data.

```{r, eval=T}
raw <- read.csv(file = "ControlMixerMSstatsInputfromskyline.csv")
head(raw)
```

The raw data (input data for MSstats) is required to contain variable of `ProteinName`, `PeptideSequence`, `PrecursorCharge`, `FragmentIon`, `ProductCharge`, `IsotopeLabelType`, `Condition`, `BioReplicate`, `Run`, `Intensity`. The variable names should be fixed. MSstats input from Skyline adapts the column scheme of the dataset so that it fits MSstats input format. However you should change `FileName` to `Run` and `Area` to `Intensity`.

@@

The difference between output from Skyline and other spectral processing tool is that Skyline can distinguish random missing (NA) and censored missing (zero). The output from Skyline can have NA (expect small number of NAs or none of them) and 0 (zero). Then, we can use zero values in intensity are censored.

@@

### 1.1 Preprocessing with DDA experiment from Skyline output

@@

### 2. Different option for dataProcess
Additional options in `dataProcess` need to be specified for processing Skyline output are `skylineReport=TRUE` (remove `Truncated=TRUE` rows and handle intensity=0) and `censoredInt="0"` (use intensity=0 as censored to handle missing values).

@@

If you have MaxQuant output, please follow this section. Here controlled mixture data with dynamic range benchmark (Cox, 2014 in MCP) is used to demonstrate.

### 1. First, get protein ID information
```{r, eval=T}
proteinGroups <- read.table("DDA2014_proteinGroups.txt", sep = "\t", header = TRUE)
```

### 2. Read in annotation including condition and biological replicates: annotation.csv
```{r, eval=T}
annot <- read.csv("DDA2014_annotation.csv", header = TRUE)
```

### 3. Read in MaxQuant file: evidence.txt
```{r, eval=F}
infile <- read.table("evidence.txt", sep = "\t", header = TRUE)
```

```{r, eval=T, echo=FALSE}
load("DDA2014.evidence.RData")
infile <- DDA2014.evidence
```

@@

```{r, eval=T}
msstats.raw <- MaxQtoMSstatsFormat(evidence=infile, annotation=annot, proteinGroups=proteinGroups)
